---
title: "proj1"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


Part 1: Step 1 - Scraping

first I loaded the url using the read_html function, then accessed the table by using its id, table-striped, within the html_node function. Then made it an html_table & set the column names to desired names, and finished the pipeline by calling the as.data.frame() function.

```{r load_and_scrape}
library(rvest)
library(magrittr)
library(tidyverse)

url <- "https://www.spaceweatherlive.com/en/solar-activity/top-50-solar-flares"

sf_tbl <- read_html(url) %>%
  html_node(".table-striped") %>%
  html_table() %>%
  set_colnames(c("rank", "flare_classification", "date", "flare_region", "start_time", "maximum_time", "end_time", "movie")) %>%
  as.data.frame()

slice(sf_tbl, 1:10)

```


Part 1: Step 2 - Tidying

First I dropped the movie column with select, then called unite on date & each of the time columns. Then I converted the new columns to POSIXct datetimes. Then dropped the old uneeded columns, start_time, end_time, max_time, and date using subset with their indexes. Then I used slice to show that the first 10 rows appear as desired.  

```{r tidy}
sf_tbl <- select(sf_tbl, -8)  %>%
  unite(start_datetime, date, start_time, sep = " ", remove = FALSE)%>%
  unite(max_datetime, date, maximum_time, sep = " ", remove = FALSE)%>%
  unite(end_datetime, date, end_time, sep = " ", remove = FALSE)
  
sf_tbl$start_datetime <- as.POSIXct(sf_tbl$start_datetime, tz = "")
sf_tbl$max_datetime <- as.POSIXct(sf_tbl$max_datetime, tz = "")
sf_tbl$end_datetime <-  as.POSIXct(sf_tbl$end_datetime, tz = "")

sf_tbl <- subset(sf_tbl, select = -c(6, 8:10))
  
slice(sf_tbl, 1:10)
```
Part 1: Step 3

In step 3, I first read in the url to the nasa data and use read_html(), html_node("body"), and html_text to store all of the nasa data as one long string in nasa_text. I then use str_split on nasa_text using the new line character as the regex to split on. I then used unlist once because it returned it as a list of lists instead of one list. after that I dropped the first 15 rows because they were not part of the desired data, as well as the last 2 rows, and then called as.data.frame() to turn it into a data frame. At this point I just needed to separate the rows into columns. I did so using separate, passing in the desired column names, and separated on whitespace. This returned the desired dataframe with an additional column that I proceeded to drop. I showed a slice of the first ten rows of my data frame. 

```{r nasa_scrape}
url2 <- "https://cdaw.gsfc.nasa.gov/CME_list/radio/waves_type2.html"

nasa_text  <- url2 %>%
  read_html() %>%
  html_node("body") %>%
  html_text()

nasa_rows = str_split(nasa_text, "\n")
nasa_rows = unlist(nasa_rows)
nasa_rows <- nasa_rows[-c(1:15, 527:529)] %>%
  as.data.frame()

nasa_table <- separate(nasa_rows, ., c("start_date", "start_time", "end_date", "end_time", "start_frequency", "end_frequency", "flare_location","flare_region", "flare_classification", "cme_date", "cme_time", "cme_angle", "cme_width", "cme_speed", "tb_deleted"), "\\s+")

nasa_table <- nasa_table[-15]

slice(nasa_table, 1:50)
```

Part 1: Step 4

For Tidying, I first gathered the indices of values matching Halo and stored the values in halos and did the same with widths to use later. I then replaced missing values with NA. After that I concatinated the year number with the end date & cme date. After I united the dates & times. Then converted to POSIX datetimes. I then matched any CME datetimes that didnt have values and replaced them with NA.I then created halo_num column & widthlimitnum colums storing halos & widths. I used those values to translate to correlating logical columns halo & cme_width. I then dropped excess columns & converted integer types to integers. 

```{r nasa_tidy}
halos <- regexpr('Halo', nasa_table$cme_angle)
widths <- regexpr('>', nasa_table$cme_width)

nasa_table[nasa_table == "—-" | nasa_table == "Halo" | nasa_table == "LASCO DATA GAP" |nasa_table == "back" | nasa_table == ""] <- NA

nasa_table$end_date <- paste(substring(nasa_table$start_date, 1,4), nasa_table$end_date, sep = "/")

nasa_table$cme_date <- paste(substring(nasa_table$start_date, 1,4), nasa_table$cme_date, sep = "/")

nasa_table <- unite(nasa_table, start_datetime, start_date, start_time, sep = " ", remove = FALSE)%>%
  unite(cme_datetime, cme_date, cme_time, sep = " ", remove = FALSE)%>%
  unite(end_datetime, end_date, end_time, sep = " ", remove = FALSE)
  
nasa_table$start_datetime <- as.POSIXct(nasa_table$start_datetime, tz = "")

nasa_table$cme_datetime <- sub("[-]+", NA, nasa_table$cme_datetime)

nasa_table$cme_datetime <- as.POSIXct(nasa_table$cme_datetime, tz = "")

nasa_table$end_datetime <- as.POSIXct(nasa_table$end_datetime, tz = "")

nasa_table$cme_width_limit_num <- widths

nasa_table$halo_num <- halos

nasa_table$halo <- FALSE
nasa_table$cme_width_limit<-FALSE

nasa_table$cme_width_limit[nasa_table$cme_width_limit_num != -1] <- TRUE

nasa_table$halo[nasa_table$halo_num != -1] <- TRUE

nasa_table <- select(nasa_table, -c(2,3,5,6,13,14,18,19))

nasa_table$start_frequency <- sapply(nasa_table$start_frequency, as.integer)

nasa_table$end_frequency <- sapply(nasa_table$end_frequency, as.integer)

nasa_table$cme_angle <- sapply(nasa_table$cme_angle, as.integer)

nasa_table$cme_speed <- sapply(nasa_table$cme_speed, as.integer)

slice(nasa_table, 1:30)

```
## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
