---
title: "Project 4"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(stringr)

```

```{r map_setup}
arrest_tab <- read_csv("http://www.hcbravo.org/IntroDataSci/misc/BPD_Arrests.csv")
dat <- arrest_tab %>% 
  filter(!is.na(`Location 1`)) %>%
  separate(`Location 1`, c("lat","lng"), sep=",") %>%
  mutate(lat=as.numeric(str_replace(lat, "\\(", ""))) %>%
  mutate(lng=as.numeric(str_replace(lng, "\\)", ""))) %>%
  sample_n(2000)
dat


```


```{r map_disp}
library(leaflet)

pal <- colorFactor(c("blue", "pink"), domain = c('M', 'F'))

dat %<>% group_by(district) %>%
  mutate(avg_lat = mean(lat)) %>%
  mutate(avg_lng = mean(lng)) %>%
  ungroup()


balto_map <- leaflet(dat) %>%
  addTiles() %>%
  setView(lat=39.29, lng=-76.61, zoom=11) %>%
  addTiles() %>%
  addCircleMarkers(popup = 
                     paste("Race:", dat$race, "<br>",
                           "Date:", dat$arrestDate ),
             clusterOptions = markerClusterOptions(),
             color = ~pal(dat$sex)) %>%
  addMarkers(lat = dat$avg_lat, lng = dat$avg_lng, popup = dat$district)

balto_map


```


Made a function pal to assign the color blue to Male arrests and pink to female arrests. Grouped the data by district and calculated the avg longitude and latitude to find a good point to add a marker for each Disctrict. District markers have a popup telling which district they are. Added circle markers and using the pal function from earlier made them pink or blue based on sex of the arrest data. Each has a popup telling the race and arrest date of the marker. Used clustering to cluster together arrests close in location. Yellow clusters indicate more data values. 


